WSClean: an implementation of a fast, generic wide-field imager

### 2014-02-17 articles/2014/Article-WSClean/wsclean.tex
Some corrections

\author[A.~R.~Offringa et al.]{A.~R.~Offringa$^{1,2}$\thanks{E-mail:
\url{andre.offringa@anu.edu.au}},
B.~McKinley\ANU$^,$\CAASTRO,
N.~Hurley-Walker\Curtin,
F.~H.~Briggs$^{1,2}$, \newauthor
R.~B.~Wayth\Curtin$^,$\CAASTRO, 
D.~L.~Kaplan\UWisc,
L.~Feng\MIT,
A.~R.~Neben\MIT,
J.~D.~Hughes\ANU, \newauthor
J.~Rhee\ANU,
G.~Bernardi\SKASA,
J.~D.~Bowman\ASU,
R.~J.~Cappallo\Haystack,
B.~E.~Corey\Haystack, \newauthor
A.~A.~Deshpande\RRI, 
D.~Emrich\Curtin,
B.~M.~Gaensler\USydney$^,$\CAASTRO,
R.~Goeke\MIT,
L.~J.~Greenhill\CfA, \newauthor
B.~J.~Hazelton\UW, 
J.~N.~Hewitt\MIT,
M.~Johnston-Hollitt\Victoria,
J.~C.~Kasper\CfA, \newauthor
E.~Kratzenberg\Haystack, 
C.~J.~Lonsdale\Haystack,
M.~J.~Lynch\Curtin,
S.~R.~McWhirter\Haystack,
D.~A.~Mitchell\CASS$^,$\CAASTRO, \newauthor
M.~F.~Morales\UW, 
E.~Morgan\MIT,
D.~Oberoi\Tata,
S.~M.~Ord\Curtin$^,$\CAASTRO,
T.~Prabu\RRI, 
A.~E.~E.~Rogers\Haystack, \newauthor
D.~A.~Roshi\NRAO \footnote{The National Radio Astronomy Observatory is a facility of the National Science Foundation operated under cooperative agreement by Associated Universities, Inc.},
N.~Udaya~Shankar\RRI,
K.~S.~Srivani\RRI,
R.~Subrahmanyan\RRI$^,$\CAASTRO, \newauthor
S.~J.~Tingay\Curtin$^,$\CAASTRO, 
M.~Waterson\Curtin$^,$\ANU,
R.~L.~Webster\UMelbourne$^,$\CAASTRO,
A.~R.~Whitney\Haystack, \newauthor
A.~Williams\Curtin,
C.~L.~Williams\MIT
\\
\ANU{}Research School of Astronomy and Astrophysics, Australian National University, Canberra, ACT 2611, Australia \\
\CAASTRO{}ARC Centre of Excellence for All-sky Astrophysics (CAASTRO), Australian National University, Canberra, ACT 2611, Australia \\
\Curtin{}International Centre for Radio Astronomy Research, Curtin University, Bentley, WA 6102, Australia\\
\UWisc{}Department of Physics, University of Wisconsin--Milwaukee, Milwaukee, WI 53201, USA\\
\MIT{}Kavli Institute for Astrophysics and Space Research, Massachusetts Institute of Technology, Cambridge, MA 02139, USA\\
\SKASA{}Square Kilometre Array South Africa (SKA SA), Cape Town 7405, South Africa\\
\ASU{}School of Earth and Space Exploration, Arizona State University, Tempe, AZ 85287, USA\\
\Haystack{}MIT Haystack Observatory, Westford, MA 01886, USA\\
\RRI{}Raman Research Institute, Bangalore 560080, India\\
\USydney{}Sydney Institute for Astronomy, School of Physics, The University of Sydney, NSW 2006, Australia\\
\CfA{}Harvard-Smithsonian Center for Astrophysics, Cambridge, MA 02138, USA\\
\UW{}Department of Physics, University of Washington, Seattle, WA 98195, USA\\
\Victoria{}School of Chemical \& Physical Sciences, Victoria University of Wellington, Wellington 6140, New Zealand\\
\CASS{}CSIRO Astronomy and Space Science, Marsfield, NSW 2122, Australia\\
\Tata{}National Centre for Radio Astrophysics, Tata Institute for Fundamental Research, Pune 411007, India\\
\NRAO{}National Radio Astronomy Observatory, Charlottesville and Greenbank, USA\\
\UMelbourne{}School of Physics, The University of Melbourne, Parkville, VIC 3010, Australia\\
}

@@

\begin{abstract}
We present a new interferometric imager that uses the $w$-stacking algorithm and can make use of the $w$-snapshot algorithm. On MWA data, we find it to be an order of magnitude faster than $w$-projection, as well as being capable of full-sky imaging at full resolution. The performance dependencies of CASA's $w$-projection and WSClean are analysed and analytical functions are derived that describe the required computing cost for both imagers. The difference in computing cost is mainly caused by the efficiency of our new imager on MWA's 27\degree field of view. We predict the computing costs for several other configurations and estimate that our imager is approximately a factor of 2--3 faster on other arrays. Imaging SKA observations is challenging, as the full SKA1 low will require 60 PetaFLOPS of computing power with current techniques. We find that the $w$-snapshot algorithm does not further improve computing requirements. The source code of our new imager is publicly released.
\end{abstract}

@@

Observations from non-coplanar interferometric radio telescopes that image large fractions of the sky at once can not be imaged with a two-dimensional Fast Fourier Transformation (FFT). Instead, the imaging algorithm needs to account for the ``$w$-term'' during inversion, which is the term that describes the deviation of the array from a perfect plane \citep{perley-noncoplanar-arrays}. This deviation is amplified for telescopes with wide field of view, making this especially an issue for low-frequency telescopes that by nature are wide-field instruments.

@@

For small field of views, the term $\sqrt{1-l^2-m^2}$ is approximately of unit size, making Eq.~\eqref{eq:visibility-function} approximate an ordinary 2-dimensional Fourier transform that can be inverted with an inverse Fast Fourier Transform (FFT). A common rule is that this is valid when
\begin{equation}\label{eq:when-2d-is-valid}
\forall w,l,m: w\left(\sqrt{1-l^2-m^2}-1\right) \ll 1.
\end{equation}

To derive the $w$-stacking technique, Eq.~\eqref{eq:visibility-function} is rewritten to
\begin{align}\notag
V(u,v,w) = & \iint \frac{I'(l,m) e^{-2\pi i w(\sqrt{1-l^2-m^2}-1)}}{\sqrt{1-l^2-m^2}} \times \\ \notag
& e^{-2\pi i \left(ul + vm\right)} dl dm.
\end{align}
This is an ordinary two-dimensional Fourier transform going from $u,v$ space to $l,m$ space, and can be inverted to get:
% Intermediate step:
%\begin{align}\notag
%\frac{I'(l,m) e^{-2\pi i w(\sqrt{1-l^2-m^2}-1)}}{\sqrt{1-l^2-m^2}} = & \iint V(u,v,w) \times \\ \notag
%& e^{-2\pi i \left(ul + vm\right)} du dv,
%\end{align}
%which can be rewritten to
\begin{align}\notag
\frac{I'(l,m)}{\sqrt{1-l^2-m^2}} = & e^{2\pi i w(\sqrt{1-l^2-m^2}-1)} \iint V(u,v,w) \times \\ \notag
& e^{2\pi i \left(ul + vm\right)} du dv.
\end{align}
Integrating both sides over $w_{\min}$ to $w_{\max}$, the minimum and maximum value of $w$, results in
\begin{align}\notag
\frac{I'(l,m)\left(w_{\max} - w_{\min}\right)}{\sqrt{1-l^2-m^2}} = \int\limits_{w_{\min}}^{w_{\max}} e^{2\pi i w(\sqrt{1-l^2-m^2}-1)} \times \\ \label{eq:wstacking}
\iint V(u,v,w)  e^{2\pi i \left(ul + vm\right)} du dv dw.
\end{align}
The final step is to make the $u,v,w$ parameters discrete, so that the integration over $u$ and $v$ can become an inverse FFT and the integration over $w$ becomes a summation. This shows that the sky function can be reconstructed by i)~grid samples with equal $w$-value on a uniform grid; ii)~calculating the inverse FFT; iii)~apply the direction dependent phaseshift $e^{2\pi i w(\sqrt{1-l^2-m^2}-1)}$; iv)~repeat this for all $w$-values and add the results together; v)~apply the final scaling.

In practice, the final scaling will be different from $\left(w_{\max} - w_{\min}\right)/\sqrt{1-l^2-m^2}$ suggested by Eq.~\eqref{eq:wstacking}, because the individual $w$-layers will not be completely filled with samples. Therefore, the number of added samples and their weight have to be taken into account. Additionally, it might be required to divide out the effect of a possible convolution kernel, the primary beam and correct for other direction dependent effects. In XX, YY, LL or RR polarizations, a correlated baseline is the complex conjugate of the reversed baseline, and the relation $V(u,v,w)=\overline{V(-u,-v,-w)}$ holds. The right-hand side of Eq.\eqref{eq:wstacking} with only positive $w$-value samples then becomes the complex conjugate of the one with only negative $w$-values samples. In this case we can therefore calculate the image for $w<0$ from the image with $w>0$. That allows us to set $w_{\min}$ to the minimum absolute $w$-value, which requires two times fewer layers. In any case, the input to the two-dimensional inverse FFT is not generally an Hermitian-symmetric function, and hence the inverse FFT is always performed as a complex-to-complex transform.

@@

\subsection{Time complexity of $w$-stacking} \label{sec:time-complexity}
It is useful to analyse the time complexity of $w$-stacking and compare it with $w$-projection, to understand which algorithm performs better in a given situation. We will use the following symbols: $N_\textrm{wlay}$ is the number of $w$-layers for $w$-stacking, $N_\textrm{pix}$ is the number of pixels in the image along each dimension, $N_\textrm{vis}$ is the number of visibilities and $N_\textrm{kern}$ is the size of the antialiasing kernel (see \S\ref{sec:gridding}), $N_\textrm{wkern}$ is the size of the $w$-kernel for $w$-projection, $w_{\max}$ is the maximum $w$ value and $\alpha_\textrm{FOV}$ is the imaging field of view.

@@

The new imager, called ``WSClean'', is written in the C++ language. It reads visibilities from CASA measurement sets and writes output images to Flexible Image Transport System (FITS) files. Several steps are multi-threaded using the threading module of the C++11 standard library. These are: reading and writing; gridding from and to different $w$-layers; performing the Fast Fourier transforms; and H\"ogbom Clean iterations (peak-finding and image subtraction, \citealt{hogbom-clean}). For the latter, intrinsics are used as well. Because of these optimizations, minor Cleaning iterations with an image size of $3072\times3072$ are performed at a rate of hundreds per second. This is fast enough to make the Clark Clean optimization \citep{clark-clean}, which consists of considering only a subset of pixels with a reduced PSF, less relevant. However, the PSF varies for imaging with non-zero $w$-values, and subtracting a constant PSF in image space leads to inaccuracies. After a number of minor iterations, it is therefore beneficial to invert the model back to the visibilities via prediction, and subtract the model directly from the visibilities \citep{wprojection-cornwell}. This is similar to the Cotton-Schwab cleaning method \citep{cotton-schwab-clean}.

@@

To efficiently image the full sky, an observation is split into short snapshots, their phase centre is changed to zenith and subsequently a suitable area is imaged. Depending on desired resolution and dynamic range, this might require very large images. To make an image at the MWA resolution, the image needs to have approximately 10K pixels in width and height. The $w$-values will be very small at zenith, because at zenith only the vertical offsets of the antennas will contribute to the $w$-value. For the MWA, the maximum height difference between tiles is $8.5$m. The MWA tiles are in fact on a slight slope, and by fitting a plane to the antennas and changing the phase centre towards the normal of that plane, the maximum $w$-value decreases to $5.5\textrm{m} / \lambda$. In the following sections, when discussing the MWA zenith direction we are referring to this optimal $w$-direction.

@@

\subsection{Beam correction}
Because the MWA consists of fixed, beamformed dipole antennas, the voltage beam of a MWA tile is described by a complex, non-diagonal Jones matrix. Alternatively, Mueller matrices can be used as well, but in general these result in more computations \citep{revisiting-me-i}. When assuming all individual antennas of the tiles are working properly, all tiles have the same beam. The beam can then be corrected in image space for all tiles at once with
\begin{equation}\label{eq:beam-correction}
 I(l, m) = J(l,m)^{-1} I'(l, m) \left(J(l,m)^*\right)^{-1},
\end{equation}
where each term is a $2\times2$ complex matrices, with $J$ the voltage beam and $^*$ denoting the conjugate transpose. Because $J$ is complex and non-diagonal, all 4 combinations of the cross-correlated polarizations $p$ and $q$\footnote{The instrumental polarizations of the MWA are informally often referred to as $X$ and $Y$, mainly because many software treats the two polarizations as such. However, this is not correct, because $X$ is defined as being parallel to the celestial equator.} including the imaginary part of the $pq$ and $qp$ are required to calculate any of the Stokes parameters. The $pp$ and $qq$ have no imaginary part due to the $uv$-symmetry. To make proper MWA beam correction possible, WSClean can output both real $pp$, $pq$, $qp$, $qq$ images and imaginary $pq$ and $qp$ images. Each polarization requires a separate run of the $w$-stacking algorithm. Once these six images are created, a separate program is used to perform the correction of Eq.~\eqref{eq:beam-correction}. This method avoids expensive beam-correcting kernels during gridding, but only works because all tiles have the same beam. To apply the same method on heterogeneous arrays such as LOFAR, each set of correlations with a different combination of station beams will have to be imaged separately, which will increase the cost of the algorithm excessively unless an a-projection kernel is used. The AWImager uses an intermediate method, and corrects a common dipole factor in image space and the phased-array beam factor in $uv$-space, which allows the gridding kernel to be smaller compared to correcting both in $uv$-space \citep{awimager-2013}.

@@

The prolate spheroidal wave function (PSWF) is generally considered to be the optimal windowing function for gridding \citep{fourier-kernel-selection-1991}. CASA convolves samples with a PSWF of 7 pixels total width during gridding. When using a variable kernel size, a PSWF is quite complicated and computational expensive to calculate. WSClean currently uses a Kaiser-Bessel (KB) window function, which is easy and fast to compute, and is a good approximation for the PSWF \citep{fourier-kernel-selection-1991}. After some benchmarks (see Appendix~\ref{sec:gridding-appendix}) we found that 7 pixels is a good compromise between performance and accuracy, and this is the default for WSClean.

@@

As can be seen in Panel~\ref{fig:timing-fov}, the field of view (FOV) has a large affect on performance when imaging off-zenith. This plot was created by varying the size of the pixel in the output image, such that the number of pixels in the image did not change. The FOV is calculated as angle between the left and right most pixel in the image. CASA is not able to make off-zenith images larger than $\sim30$\degree, and its imaging cost increases much more rapidly compared to the cost of WSClean, which implies that gridding is the major cost in this scenario. The image recentering technique is clearly beneficial for FOVs larger than $\sim30$\degree, and can even make a difference of an order of magnitude at very large FOVs of $\sim90$\degree.

@@

\subsection{Optimal snapshot duration}
TODO: need to recalculate this with proper FOV and ZA corrections.
Snapshot imaging, such as the recentring technique described in \S\ref{sec:snapshot-imaging}, can be implemented with either $w$-projection or $w$-stacking. Using the derived formula, we can estimate the cost of making snapshots with both techniques. Snapshot imaging effectively removes the dependency on ZA from $w_{\max}$. Given the snapshot duration $\Delta \tau_\textrm{snapshot}$ and the total integration time $\Delta \tau_\textrm{total}$, the total time cost of inversion becomes
\begin{equation} \label{eq:snapshot-cost}
t_\textrm{snapshot}(\Delta \tau_\textrm{snapshot}) = \frac{\Delta \tau_\textrm{total}}{\Delta \tau_\textrm{snapshot}} t_\textrm{imaging}(w'_{\max},N'_\textrm{vis},N_\textrm{freq},N_\textrm{pix}),
\end{equation}
with
\begin{equation}
w'_{\max} = \frac{1}{\lambda}\max \Delta z \left(1.0 - \cos(\frac{1}{2}(\textrm{FOV} + \omega_E \tau_\textrm{snapshot})\right),
\end{equation}
$\omega_E$ the rotational speed of the Earth and $N'_\textrm{vis} = N_\textrm{vis}\frac{\Delta \tau_\textrm{total}}{\tau_\textrm{snapshot}}$. We have excluded the cost of phase shifting the visibilities and gridding. The cost for regridding with simple nearest neighbour or bilinair interpolation is indeed negligable, although more accurate interpolation (e.g. Lanczos interpolation) can be expensive. Also, our current implementation of the phase-changing program does take a non-negligable time, but this implementation is not optimized and can in theory be implemented in a preprocessing pipeline or on the fly during imaging.

>>>>

\author[A.~R.~Offringa et al.]{A.~R.~Offringa$^{1,2}$\thanks{\hspace{2cm}E-mail: \url{andre.offringa@anu.edu.au}},
B.~McKinley\ANU$^,$\CAASTRO,
N.~Hurley-Walker\Curtin,
F.~H.~Briggs$^{1,2}$, \newauthor
R.~B.~Wayth\Curtin$^,$\CAASTRO, 
D.~L.~Kaplan\UWisc,
L.~Feng\MIT,
A.~R.~Neben\MIT,
J.~D.~Hughes\ANU, \newauthor
J.~Rhee\ANU,
G.~Bernardi\SKASA,
J.~D.~Bowman\ASU,
R.~J.~Cappallo\Haystack,
B.~E.~Corey\Haystack, \newauthor
A.~A.~Deshpande\RRI, 
D.~Emrich\Curtin,
B.~M.~Gaensler\USydney$^,$\CAASTRO,
R.~Goeke\MIT,
L.~J.~Greenhill\CfA, \newauthor
B.~J.~Hazelton\UW, 
J.~N.~Hewitt\MIT,
M.~Johnston-Hollitt\Victoria,
J.~C.~Kasper\CfA, \newauthor
E.~Kratzenberg\Haystack, 
C.~J.~Lonsdale\Haystack,
M.~J.~Lynch\Curtin,
S.~R.~McWhirter\Haystack,
D.~A.~Mitchell\CASS$^,$\CAASTRO, \newauthor
M.~F.~Morales\UW, 
E.~Morgan\MIT,
D.~Oberoi\Tata,
S.~M.~Ord\Curtin$^,$\CAASTRO,
T.~Prabu\RRI, 
A.~E.~E.~Rogers\Haystack, \newauthor
D.~A.~Roshi\NRAO \footnote{The National Radio Astronomy Observatory is a facility of the National Science Foundation operated under cooperative agreement by Associated Universities, Inc.},
N.~Udaya~Shankar\RRI,
K.~S.~Srivani\RRI,
R.~Subrahmanyan\RRI$^,$\CAASTRO, \newauthor
S.~J.~Tingay\Curtin$^,$\CAASTRO, 
M.~Waterson\Curtin$^,$\ANU,
R.~L.~Webster\UMelbourne$^,$\CAASTRO,
A.~R.~Whitney\Haystack, \newauthor
A.~Williams\Curtin,
C.~L.~Williams\MIT
\\
\ANU{}Research School of Astronomy and Astrophysics, Australian National University, Canberra, ACT 2611, Australia \\
\CAASTRO{}ARC Centre of Excellence for All-sky Astrophysics (CAASTRO), Australian National University, Canberra, ACT 2611, Australia \\
\Curtin{}International Centre for Radio Astronomy Research, Curtin University, Bentley, WA 6102, Australia\\
\UWisc{}Department of Physics, University of Wisconsin--Milwaukee, Milwaukee, WI 53201, USA\\
\MIT{}Kavli Institute for Astrophysics and Space Research, Massachusetts Institute of Technology, Cambridge, MA 02139, USA\\
\SKASA{}Square Kilometre Array South Africa (SKA SA), Cape Town 7405, South Africa\\
\ASU{}School of Earth and Space Exploration, Arizona State University, Tempe, AZ 85287, USA\\
\Haystack{}MIT Haystack Observatory, Westford, MA 01886, USA\\
\RRI{}Raman Research Institute, Bangalore 560080, India\\
\USydney{}Sydney Institute for Astronomy, School of Physics, The University of Sydney, NSW 2006, Australia\\
\CfA{}Harvard-Smithsonian Center for Astrophysics, Cambridge, MA 02138, USA\\
\UW{}Department of Physics, University of Washington, Seattle, WA 98195, USA\\
\Victoria{}School of Chemical \& Physical Sciences, Victoria University of Wellington, Wellington 6140, New Zealand\\
\CASS{}CSIRO Astronomy and Space Science, Marsfield, NSW 2122, Australia\\
\Tata{}National Centre for Radio Astrophysics, Tata Institute for Fundamental Research, Pune 411007, India\\
\NRAO{}National Radio Astronomy Observatory, Charlottesville and Greenbank, USA\\
\UMelbourne{}School of Physics, The University of Melbourne, Parkville, VIC 3010, Australia\\
}

@@

\begin{abstract}
We present a new interferometric imager that uses the $w$-stacking algorithm and can make use of the $w$-snapshot algorithm. On MWA data, we find it to be an order of magnitude faster than $w$-projection, as well as being capable of full-sky imaging at full resolution and with full polarization correction. The performance dependencies of CASA's $w$-projection and WSClean are analysed and analytical functions are derived that describe the required computing cost for both imagers. The difference in computing cost is mainly caused by the efficiency of our new imager on MWA's 27\degree field of view. We predict the computing costs for several other configurations and estimate that our imager is approximately a factor of 2--3 faster on other arrays. Imaging SKA observations is challenging, as we estimate the computing cost of full SKA1 low imaging to be 60 PetaFLOPS with current techniques. We find that the $w$-snapshot algorithm does not further improve computing requirements. The source code of our new imager is publicly released.
\end{abstract}

@@

Observations from non-coplanar interferometric radio telescopes that image large fractions of the sky at once can not be imaged with a two-dimensional fast Fourier transform (FFT). Instead, the imaging algorithm needs to account for the ``$w$-term'' during inversion, which is the term that describes the deviation of the array from a perfect plane \citep{perley-noncoplanar-arrays}. This deviation is amplified for telescopes with wide field of view, making this especially an issue for low-frequency telescopes that by nature are wide-field instruments.

@@

For small field of views, the term $\sqrt{1-l^2-m^2}$ is approximately of unit size, making Eq.~\eqref{eq:visibility-function} approximate an ordinary 2-dimensional Fourier transform that can be inverted with an inverse FFT. A common rule is that this is valid when
\begin{equation}\label{eq:when-2d-is-valid}
\forall w,l,m: w\left(\sqrt{1-l^2-m^2}-1\right) \ll 1.
\end{equation}

To derive the $w$-stacking technique, Eq.~\eqref{eq:visibility-function} is rewritten to
\begin{align}\notag
V(u,v,w) = & \iint \frac{I'(l,m) e^{-2\pi i w(\sqrt{1-l^2-m^2}-1)}}{\sqrt{1-l^2-m^2}} \times \\ \notag
& e^{-2\pi i \left(ul + vm\right)} dl dm.
\end{align}
This is an ordinary two-dimensional Fourier transform going from $u,v$ space to $l,m$ space, and can be inverted to get:
% Intermediate step:
%\begin{align}\notag
%\frac{I'(l,m) e^{-2\pi i w(\sqrt{1-l^2-m^2}-1)}}{\sqrt{1-l^2-m^2}} = & \iint V(u,v,w) \times \\ \notag
%& e^{-2\pi i \left(ul + vm\right)} du dv,
%\end{align}
%which can be rewritten to
\begin{align}\notag
\frac{I'(l,m)}{\sqrt{1-l^2-m^2}} = & e^{2\pi i w(\sqrt{1-l^2-m^2}-1)} \iint V(u,v,w) \times \\ \notag
& e^{2\pi i \left(ul + vm\right)} du dv.
\end{align}
Integrating both sides over $w_{\min}$ to $w_{\max}$, the minimum and maximum value of $w$, results in
\begin{align}\notag
\frac{I'(l,m)\left(w_{\max} - w_{\min}\right)}{\sqrt{1-l^2-m^2}} = \int\limits_{w_{\min}}^{w_{\max}} e^{2\pi i w(\sqrt{1-l^2-m^2}-1)} \times \\ \label{eq:wstacking}
\iint V(u,v,w)  e^{2\pi i \left(ul + vm\right)} du dv dw.
\end{align}
The final step is to make the $u,v,w$ parameters discrete, so that the integration over $u$ and $v$ can become an inverse FFT and the integration over $w$ becomes a summation. This shows that the sky function can be reconstructed by i)~grid samples with equal $w$-value on a uniform grid; ii)~calculating the inverse FFT; iii)~apply the direction-dependent phaseshift $e^{2\pi i w(\sqrt{1-l^2-m^2}-1)}$; iv)~repeat this for all $w$-values and add the results together; v)~apply the final scaling.

In practice, the final scaling will be different from $\left(w_{\max} - w_{\min}\right)/\sqrt{1-l^2-m^2}$ suggested by Eq.~\eqref{eq:wstacking}, because the individual $w$-layers will not be completely filled with samples. Therefore, the number of added samples and their weight have to be taken into account. Additionally, it might be required to divide out the effect of a possible convolution kernel, the primary beam and correct for other direction-dependent effects. In XX, YY, LL or RR polarizations, a correlated baseline is the complex conjugate of the reversed baseline, and the relation $V(u,v,w)=\overline{V(-u,-v,-w)}$ holds. The right-hand side of Eq.\eqref{eq:wstacking} with only positive $w$-value samples then becomes the complex conjugate of the one with only negative $w$-values samples. In this case we can therefore calculate the image for $w<0$ from the image with $w>0$. That allows us to set $w_{\min}$ to the minimum absolute $w$-value, which requires two times fewer layers. In any case, the input to the two-dimensional inverse FFT is not generally an Hermitian-symmetric function, and hence the inverse FFT is always performed as a complex-to-complex transform.

@@

\subsection{Time complexity of $w$-stacking} \label{sec:time-complexity}
It is useful to analyse the time complexity of $w$-stacking and compare it with $w$-projection, to understand which algorithm performs better in a given situation. We will use the following symbols: $N_\textrm{wlay}$ is the number of $w$-layers for $w$-stacking, $N_\textrm{pix}$ is the number of pixels in the image along each dimension, $N_\textrm{vis}$ is the number of visibilities, $N_\textrm{kern}$ is the size of the antialiasing kernel (see \S\ref{sec:gridding}), $N_\textrm{wkern}$ is the size of the $w$-kernel for $w$-projection, $w_{\max}$ is the maximum $w$ value and $\alpha_\textrm{FOV}$ is the imaging field of view.

@@

The new imager, called ``WSClean'', is written in the C++ language. It reads visibilities from CASA measurement sets and writes output images to Flexible Image Transport System (FITS) files. Several steps are multi-threaded using the threading module of the C++11 standard library. These are: reading and writing; gridding from and to different $w$-layers; performing the FFTs; and H\"ogbom Clean iterations (peak-finding and image subtraction, \citealt{hogbom-clean}). For the latter, intrinsics are used as well. Because of these optimizations, minor Cleaning iterations with an image size of $3072\times3072$ are performed at a rate of hundreds per second. This is fast enough to make the Clark Clean optimization \citep{clark-clean}, which consists of considering only a subset of pixels with a reduced PSF, less relevant. However, the PSF varies for imaging with non-zero $w$-values, and subtracting a constant PSF in image space leads to inaccuracies. After a number of minor iterations, it is therefore beneficial to invert the model back to the visibilities via prediction, and subtract the model directly from the visibilities \citep{wprojection-cornwell}. This is similar to the Cotton-Schwab cleaning method \citep{cotton-schwab-clean}.

@@

To efficiently image the full sky, an observation is split into short snapshots, their phase centre is changed to zenith and subsequently a suitable area is imaged. Depending on desired resolution and dynamic range, this might require very large images. To make an image at the MWA resolution, the image needs to have approximately 10k pixels in width and height. The $w$-values will be very small at zenith, because at zenith only the vertical offsets of the antennas will contribute to the $w$-value. For the MWA, the maximum height difference between tiles is $8.5$m. The MWA tiles are in fact on a slight slope, and by fitting a plane to the antennas and changing the phase centre towards the normal of that plane, the maximum $w$-value decreases to $5.5\textrm{m} / \lambda$. In the following sections, when discussing the MWA zenith direction we are referring to this optimal $w$-direction.

TODO: add big-ass full-sky image

@@

TODO: I just realized that the original definition of $w$-snapshots might be Natasha's way of fixing $w$-values in MWACS. In that case, these methods are new and it might be good to distinguish this.

\subsection{Beam correction}
Because the MWA consists of fixed, beamformed dipole antennas, the voltage beam of a MWA tile is described by a complex, non-diagonal Jones matrix. Alternatively, Mueller matrices can be used as well, but in general these result in more computations \citep{revisiting-me-i}. When assuming all individual antennas of the tiles are working properly, all tiles have the same beam. The beam can then be corrected in image space for all tiles at once with
\begin{equation}\label{eq:beam-correction}
 I(l, m) = J(l,m)^{-1} I'(l, m) \left(J(l,m)^*\right)^{-1},
\end{equation}
where each term is a $2\times2$ complex matrices, with $J$ the voltage beam and $^*$ denoting the conjugate transpose. Because $J$ is complex and non-diagonal, all 4 combinations of the cross-correlated polarizations $p$ and $q$\footnote{The instrumental polarizations of the MWA are informally often referred to as $X$ and $Y$, mainly because many software treats the two polarizations as such. However, this is incorrect because i) $X$ is normally defined as being parallel to the horizon; and ii) ($2\times$) Stokes I$\neq pp + qq$.} including the imaginary part of the $pq$ and $qp$ are required to calculate any of the Stokes parameters. The $pp$ and $qq$ have no imaginary part due to the $uv$-symmetry. To make proper MWA beam correction possible, WSClean can output both real $pp$, $pq$, $qp$, $qq$ images and imaginary $pq$ and $qp$ images. Each polarization requires a separate run of the $w$-stacking algorithm. Once these six images are created, a separate program is used to perform the correction of Eq.~\eqref{eq:beam-correction}. This method avoids expensive beam-correcting kernels during gridding, but only works because all tiles have the same beam. To apply the same method on heterogeneous arrays such as LOFAR, each set of correlations with a different combination of station beams will have to be imaged separately, which will increase the cost of the algorithm excessively unless an a-projection kernel is used. The AWImager uses an intermediate method, and corrects a common dipole factor in image space and the phased-array beam factor in $uv$-space, which allows the gridding kernel to be smaller compared to correcting both in $uv$-space \citep{awimager-2013}.

TODO: add Stokes I and V images of polarized source (e.g. Martin's source, if he agrees)

@@

The prolate spheroidal wave function (PSWF) is generally considered to be the optimal windowing function for gridding \citep{fourier-kernel-selection-1991}. CASA convolves samples with a PSWF of 7 pixels total width during gridding. When using a variable kernel size, a PSWF is quite complicated and computational expensive to calculate. WSClean currently uses a Kaiser-Bessel (KB) window function, which is easy and fast to compute, and is a good approximation of the PSWF \citep{fourier-kernel-selection-1991}. After some benchmarks (see Appendix~\ref{sec:gridding-appendix}) we found that 7 pixels is a good compromise between performance and accuracy, and this is the default for WSClean.

@@

As can be seen in Panel~\ref{fig:timing-fov}, the field of view (FOV) has a large affect on performance when imaging off-zenith. This plot was created by varying the size of a pixel in the output image, such that the number of pixels in the image did not change. The FOV is calculated as angle between the left- and right-most pixel in the image. CASA is not able to make off-zenith images larger than $\sim$30\degree, and its imaging cost increases much more rapidly compared to the cost of WSClean, which implies that gridding is the major cost in this scenario. The image recentering technique is clearly beneficial for FOVs larger than $\sim$30\degree, and can even make a difference of an order of magnitude at very large FOVs of $\sim$90\degree.

@@

\subsection{Optimal snapshot duration}
Snapshot imaging, such as the recentring technique described in \S\ref{sec:snapshot-imaging}, can be implemented with either $w$-projection or $w$-stacking. Using the derived formulae, we can estimate the cost of making snapshots with both techniques. Snapshot imaging effectively removes the dependency on ZA from $w_{\max}$. Given the snapshot duration $\Delta \tau_\textrm{snapshot}$ and the total integration time $\Delta \tau_\textrm{total}$, the total time cost of inversion becomes
\begin{equation} \label{eq:snapshot-cost}
t_\textrm{snapshot}(\Delta \tau_\textrm{snapshot}) = \frac{\Delta \tau_\textrm{total}}{\Delta \tau_\textrm{snapshot}} t_\textrm{imaging}(w'_{\max},N'_\textrm{vis},N_\textrm{freq},N_\textrm{pix}),
\end{equation}
with
\begin{equation}
w'_{\max} = \frac{1}{\lambda}\max \Delta z \left(1.0 - \cos(\frac{1}{2}(\textrm{FOV} + \omega_E \tau_\textrm{snapshot})\right),
\end{equation}
$\omega_E$ the rotational speed of the Earth and $N'_\textrm{vis} = N_\textrm{vis}\frac{\Delta \tau_\textrm{total}}{\tau_\textrm{snapshot}}$. We have excluded the cost of phase shifting the visibilities and gridding. The cost for regridding with simple nearest neighbour or bilinair interpolation is indeed negligable, although more accurate interpolation (e.g. Lanczos interpolation) can be expensive. Also, our current implementation of the phase-changing program does take a non-negligable time, but this implementation is not optimized and can in theory be implemented in a preprocessing pipeline or on the fly during imaging.
